Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-12-09 bsis_paper/paper.tex
Another editing pass. Changes in all sections. Updated bibliography file for articles published recently.
git-svn-id: file:///home/rblake/svn/pamg/trunk@40 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\begin{abstract}
Independent set-based coarse-grid selection algorithms are defined by
policies for how to do weight initialization, independent set
selection, and weight update. In this paper, we develop theory
demonstrating that all algorithms employing identical policies produce
identical coarse grids, regardless of the implementation. This result
on coarse grid invariance is applied to produce a new coarse-grid
selection algorithm that is substantially more efficient than an
existing algorithm using the same policies. Experimental results
comparing two versions of the new algorithm and the older algorithm
are presented, followed by a discussion on possible approaches to
effectively parallelize the new methods.
\end{abstract}

@@

AMG algorithms execute in two phases: the setup phase and the solve
phase. Setup phase algorithms provide functionality for constructing
elements needed by the solve phase, such as coarse grids, operators to
transfer information between levels of the problem (i.e., restriction
and prolongation operators), and the linear systems to be solved on
each level. The solve phase applies inexpensive iterative methods,
called relaxation methods in the context of AMG, on the problems
generated by the setup phase and uses the transfer operators to move
vectors between levels.

In this paper, we focus attention on coarse-grid selection algorithms
in the AMG setup phase. The independent set-based class of coarsening
algorithms is studied in detail. In Section~\ref{sec:cgs}, we discuss
current coarse-grid selection algorithms. The components of
independent set-based coarsening algorithms are
discussed. Section~\ref{sec:invariance} contains theory based on
general independent set-based coarsening and the invariance of coarse
grids selected when the same selection policies are
used. Section~\ref{sec:bsis} introduces a new coarsening algorithm,
Bucket-Sorted Independent Sets (BSIS), that, from the theory in
Section~\ref{sec:invariance}, produces the same coarse grids as
another coarsening algorithm in a much more efficient
manner. Experimental results are reported to demonstrate the
performance gains in BSIS discussion is included to propose a method
for parallelization of BSIS. Conclusions are presented in
Section~\ref{sec:conclusions}.

\section{Coarse-Grid Selection}
\label{sec:cgs}
Coarse-grid selection is the process of selecting the degrees of
freedom for the coarse-level problem. Classical AMG uses nodal
coarsening in which coarse-level degrees of freedom are a subset of
the degrees of freedom on finer levels. Smoothed
aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
forms a coarse-level degree of freedom by aggregating several
fine-level degrees of freedom. In this paper, we study nodal
coarsening.

Given a set of fine-level degrees of freedom, $\Omega_k$, nodal
coarse-grid selection constructs a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed for selecting the coarse grid, with independent set-based
approaches being the most numerous. Independent set-based algorithms
trace their roots to the classical AMG coarsening algorithm,
Ruge-St\"uben (RS) coarsening~\cite{Ruge1987}, also called classical
coarsening. These methods are our focus and are discussed in more
detail following a brief overview of other techniques used for nodal
coarsening.

Other methods take substantially different approaches to
coarsening. The technique in~\cite{Griebel2006} develops a parallel
coarsening algorithm that utilizes RS. On each processor, RS selects
several different coarse grids, and then one coarse grid is selected
per processor in a way to minimize special treatment on processor
boundaries. In~\cite{MacLachlan2006}, a greedy approach is used to
produce a good splitting of the unknowns into fine-grid and
coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
offer another approach for parallel coarse-grid selection by
decoupling coarse grids and alleviating the need for communication on
coarse levels. Compatible relaxation
approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
incrementally and decide when the coarse grid is sufficient through a
quality measure based on relaxation.

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
Many independent set-based coarsening algorithms form $\Omega_{k+1}$
using the following heuristics.
\begin{description}
\item[H1:] For each unknown $j$ that strongly influences $F$-point
  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
  $k$ that also strongly influences $i$.
\item[H2:] The set of $C$-points needs to form a maximal independent
  set in the reduced graph of $S$ such that no $C$-point strongly
  depends on another $C$-point.
\end{description}
H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
enforced by independent set-based methods, while H2 serves as a
guideline whose purpose is to encourage selection of small
coarse-grid point sets (i.e., small $|\Omega_{k+1}|$).

Certain independent set-based coarsening algorithms use a variant of
H1, called \hp, which is designed to lead to sparser coarse-level
matrices.
\begin{description}
\item[\hp] Each $F$-point must be strongly influenced by at least one
  $C$-point.
\end{description}

Independent set-based coarse-grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is shown in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $C \leftarrow C \cup D$
   $F \leftarrow F \setminus D$
   $(\Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(D,\, \Omega,\, w)$%
end
\end{lstlisting}

Each vertex in the graph is assigned a weight based on its merit to
becoming a $C$-point. The largest weights are conferred to vertices
best suited to be $C$-points, where the quality of potential
$C$-points is quantified by \emph{strength of connection} measures. In
the classical definition, the set of unknowns that unknown $i$
strongly depends upon is based on the relative sizes of off-diagonal
nonzeros in row $i$ of $A$ and is defined as
\begin{equation}
\label{2:eqn:S_i}
S_i = \left\{j: i \ne j \textrm{ and }|a_{ij}| \ge \theta \max_{k \ne
i}|a_{ik}|\right\},
\end{equation}
where $a_{ij}$ is the nonzero in row $i$, column $j$ of matrix $A$ and
$0 < \theta \le 1$. The set of unknowns that $i$ strongly influences,
denoted $S_i^T$, is defined as the set of unknowns that strongly
depend on $i$: $S_i^T = \{j: i \in S_j\}$. The strong influence and
strong dependence sets are used in the definition of the
\emph{strength matrix} $S$, where the entry in row $i$, column $j$ of
$S$ is defined as
\begin{equation}
S_{ij} = \left\{
\begin{array}{ll}
  1 & \textrm{if } j \in S_i, \\
  0 & \textrm{otherwise}.
\end{array}\right.
\end{equation}

@@

The selection of the independent set $D$ works by comparing weights in
each node's neighborhood, which is defined as follows.
\begin{defn}
\label{def:neighborhood}
The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
union of the sets containing the strong influences of $i$ and the
strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
\end{defn}

@@

The weight update routines in independent set-based coarsening
algorithms are varied but take one of two forms: routines that
decrease the values of weights and routines that increase the value of
weights. In all cases, a vertex with a large weight has high fitness
for becoming a $C$-point.

@@

Algorithms relying on different and larger neighborhoods, such as
\emph{distance-$d$ neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

@@

When a vertex $i$ satisfies the generalized selection
condition~(\ref{eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

@@

Let $D_0$ be the set of all vertices satisfying
(\ref{eqn:general-selection}) in the first iteration of coarse-grid
selection. Vertices in $D_0$ become $C$-points regardless of the
algorithm used to build coarse grids. In most independent set-based
algorithms developed so far, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

Lemma~\ref{5:lemma:D0_invariant} states that any coarse-grid selection
method using the general selection condition invariably selects $D_0$
vertices as $C$-points. This result is used in the proof of the next
theorem.

\begin{thm}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{eqn:general-selection}), and weight
update heuristics as described above select identical coarse grids.
\end{thm}
\begin{proof}
  Let $c$ be the vertices in the extended selection neighborhood of
  $i$ (i.e., $\mathcal{N}_i^{2s}$) satisfying the conditions for $D$
  in some arbitrary iteration, where $i \notin c$. Suppose assigning
  vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$
  for all $j \in \mathcal{N}_i^s$. Additionally, suppose assigning
  $c_2 \subset c$, $c_1 \ne c_2$, to the $C$-point set leads to $w_j >
  w_k$ for all $k \in \mathcal{N}_j^s$ for some $j \in
  \mathcal{N}_i^s$, $j \notin c$. In the first case, $i$ becomes a
  $C$-point before $j$, and the opposite is true in the second case.

@@

Both cases are impossible, so the order of $C$-point selection within
the selection neighborhood of each vertex is invariant. Combined with
Lemma~\ref{5:lemma:D0_invariant} as the base case, this proves by
induction the invariance of coarse-grid selection for all algorithms
using identical selection conditions.
\end{proof}

@@

\section{Bucket-Sorted Independent Set Selection}
\label{sec:bsis}
In this section, we introduce Bucket-Sorted Independent Sets
(BSIS). BSIS uses the same policies as CLJP-c, so from the theory in
Section~\ref{sec:invariance}, BSIS and CLJP-c produce the same coarse
grids. BSIS is designed to be more efficient than CLJP-c.

\subsection{CLJP-c}
CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
\label{cor:same-weight-indep-set}
Any set of vertices in the graph of $S$ sharing the same weight are an
independent set.
\end{cor}
\begin{cor}
\label{cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~(\ref{eqn:general-selection}). That is,
each vertex in that independent set has a uniquely maximal weight in
its neighborhood.
\end{cor}
Corollary~\ref{cor:same-weight-indep-set} states that independent sets
can be selected in the graph simply by selecting sets of vertices with
the same weight. Corollary~\ref{cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight in the
graph, updating weights, selecting the next set of vertices with
maximum weight, and so on. This approach is taken by the BSIS
algorithm.

@@

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
%\textsc{Search-Graph}$(S,\, C,\, F)$%
$D \leftarrow \emptyset$
for $i \notin (C \cup F)$
  if $w_i > w_j\ \textrm{for all } j \in \mathcal{N}_i$
    $D \leftarrow D \cup \{i\}$
  end
end
return $D$
\end{lstlisting}
% \begin{algorithm}[h]
% \caption{Coarse-Grid Selection Graph Search}
% \label{5:alg:search}
% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
% \begin{algorithmic}[1]
% \STATE $D \leftarrow \emptyset$
% \FORALL{$i \notin (C \cup F)$}
% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
% \STATE $D \leftarrow D \cup \{i\}$
% \ENDIF
% \ENDFOR
% \RETURN $D$
% \end{algorithmic}
% \}
% \end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

@@

Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
routines for search and weight update. Furthermore, rather than
applying the color information to augment vertex weights, BSIS uses
the colors in a bucket data structure. Once initialized, this data
structure selects independent sets, which satisfy the conditions in
(\ref{5:eqn:D-conditions}), in constant time.

@@

Timing data for the selection of all coarse grids in the hierarchy is
reported. This time includes the cost for all parts of the algorithms,
including the graph coloring phase. AMG solve phase information is not
reported since the algorithms produce identical coarse grids and since
information on solve phase performance for AMG with CLJP-c is
documented in~\cite{alber-cljpc,alber-PCGS}.

The smallest problem is a $30 \times 30 \times 30$ grid. Subsequent
problems are grids of size $60^3$, $90^3$, up to $210^3$. The largest
problem is 343 times larger than the smallest problem and contains
more than nine million degrees of freedom.

>>>>

\begin{abstract}
Independent set-based coarse-grid selection algorithms are defined by
policies for weight initialization, independent set selection, and
weight update. In this paper, we develop theory demonstrating that
algorithms employing the same policies produce identical coarse grids,
regardless of the implementation. This result on coarse grid
invariance is applied to produce a new coarse-grid selection algorithm
that is substantially more efficient than an existing algorithm using
the same policies. Experimental results comparing two versions of the
new algorithm and the older algorithm are presented, followed by a
discussion on possible approaches to effectively parallelize the new
method.
\end{abstract}

@@

AMG algorithms execute in two phases: the setup phase and the solve
phase. Setup phase algorithms provide functions for constructing
elements needed by the solve phase, such as coarse grids, operators to
transfer information between levels of the problem (i.e., restriction
and prolongation operators), and the linear systems to be solved on
each level. The solve phase applies inexpensive iterative methods,
called relaxation methods in the context of AMG, on the problems
generated by the setup phase and uses the transfer operators to move
vectors between levels.

In this paper, we focus attention on coarse-grid selection algorithms
in the AMG setup phase. The independent set-based class of coarsening
algorithms is studied in detail. New theory is developed to
demonstrate invariance in the result of independent set-based coarse
grid selection, given restrictions on the selection
policies. Algorithms that fall within the assumptions of the theory
are guaranteed to produce the same coarse grids, regardless of how the
policies are implemented. A new coarse-grid selection algorithm,
designed to select coarse grids more efficiently than the CLJP-c
algorithm, is presented to demonstrate the theoretical results. A
particularly attractive result of the theory is that computational
efficiency concerns become decoupled from the output of coarse-grid
selection.

The remainder of this paper is organized as follows. In
Section~\ref{sec:cgs}, we discuss contemporary coarse-grid selection
algorithms. The components of independent set-based coarsening
algorithms are presented. Section~\ref{sec:invariance} develops theory
based on general independent set-based coarsening and the invariance
of coarse grids selected when the same selection policies are
used. Section~\ref{sec:bsis} demonstrates the results from
Section~\ref{sec:invariance} through the introduction of a new
coarsening algorithm called Bucket-Sorted Independent Sets
(BSIS). BSIS produces the same coarse grids as the CLJP-c coarsening
algorithm, but is designed to do so in a much more efficient
manner. Experimental results are reported to demonstrate the
performance gains in BSIS, and discussion is included to propose a
method for the parallelization of BSIS. Conclusions are presented in
Section~\ref{sec:conclusions}.

\section{Coarse-Grid Selection}
\label{sec:cgs}
Coarse-grid selection algorithms construct the degrees of freedom for
the coarse-level problems. Classical AMG implements nodal coarsening
in which coarse-level degrees of freedom form a subset of the degrees
of freedom on finer levels. Smoothed
aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
forms a coarse-level degree of freedom by aggregating several
fine-level degrees of freedom. In this paper, we study nodal
coarsening.

Given a set of fine-level degrees of freedom, $\Omega_k$, nodal
coarse-grid selection constructs a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed to select the coarse grid, with independent set-based
approaches being the most numerous. Independent set-based algorithms
descend from the classical AMG coarsening algorithm, Ruge-St\"uben
(RS) coarsening~\cite{Ruge1987}, also called classical
coarsening. Independent set-based methods are our focus and are
discussed in more detail in the next section.

Other methods take substantially different approaches to
coarsening. The technique in~\cite{Griebel2006} develops a parallel
coarsening algorithm that utilizes RS. On each processor, RS selects
several different coarse grids, and then one coarse grid is selected
per processor in a way to minimize special treatment on processor
boundaries. In~\cite{MacLachlan2006}, a greedy approach is used to
produce a good splitting of the unknowns into fine-grid and
coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
offer another approach for parallel coarse-grid selection by
decoupling coarse grids and alleviating the need for communication on
coarse levels. Compatible relaxation
approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
incrementally and determine when the coarse grid is sufficient through
a quality measure based on relaxation.

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
Independent set-based coarsening algorithms rely on a \emph{strength
  of connection} measure to estimate the level of correlation between
  degrees of freedom. In the classical definition, the set of unknowns
  that unknown $i$ strongly depends on is based on the relative sizes
  of off-diagonal nonzeros in row $i$ of $A$ and is defined as
\begin{equation}
\label{2:eqn:S_i}
S_i = \left\{j: i \ne j \textrm{ and }|a_{ij}| \ge \theta \max_{k \ne
i}|a_{ik}|\right\},
\end{equation}
where $a_{ij}$ is the nonzero in row $i$, column $j$ of matrix $A$ and
$0 < \theta \le 1$. The set of unknowns that $i$ strongly influences,
denoted $S_i^T$, is defined as the set of unknowns that strongly
depend on $i$: $S_i^T = \{j: i \in S_j\}$. The strong influence and
strong dependence sets are used in the definition of the
\emph{strength matrix} $S$, where the entry in row $i$, column $j$ of
$S$ is defined as
\begin{equation}
S_{ij} = \left\{
\begin{array}{ll}
  1 & \textrm{if } j \in S_i, \\
  0 & \textrm{otherwise}.
\end{array}\right.
\end{equation}

@@

Many independent set-based coarsening algorithms use strong influence
and strong dependence to form the set of coarse-grid points
$\Omega_{k+1}$ using the following heuristics.
\begin{description}
\item[H1:] For each unknown $j$ that strongly influences $F$-point
  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
  $k$ that also strongly influences $i$.
\item[H2:] The set of $C$-points should form a maximal independent set
  in the reduced graph of $S$ such that no $C$-point strongly depends
  on another $C$-point.
\end{description}
H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
enforced by independent set-based methods, while H2 serves as a
guideline whose purpose is to encourage selection of small
coarse-grid point sets (i.e., small $|\Omega_{k+1}|$).

Certain independent set-based coarsening algorithms use a variant of
H1, called \hp, which is designed to produce sparser coarse-level
matrices.
\begin{description}
\item[\hp] Each $F$-point must be strongly influenced by at least one
  $C$-point.
\end{description}

Independent set-based coarse-grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is outlined in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $C \leftarrow C \cup D$
   $(F,\, w) \leftarrow\,$%\textsc{Update-Weights}$(D,\, F,\, \Omega,\, w)$%
end
\end{lstlisting}

Each vertex in the graph is assigned a weight based on its merit for
becoming a $C$-point. The largest weights are conferred to vertices
best suited to be $C$-points, where the quality of potential
$C$-points is quantified by strength of connection measures.

The selection of the independent set $D$ in Line~6 of
Algorithm~\ref{alg:general-indep-set} works by comparing weights in
each node's neighborhood, which is defined as follows.
\begin{defn}
\label{def:neighborhood}
The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
union of the sets containing the strong influences of $i$ and the
strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
\end{defn}

@@

The weight update routines in independent set-based coarsening
algorithms (encapsulated in Line~8 of
Algorithm~\ref{alg:general-indep-set}) are varied but take one of two
forms: routines that decrease the values of weights and routines that
increase the value of weights. In all cases, a vertex with a large
weight has high fitness and is more desirable as a $C$-point. As the
pseudocode is formatted in Algorithm~\ref{alg:general-indep-set},
assignment of vertices to the $F$-point set occurs during the weight
update routine.

@@

Algorithms relying on different and larger neighborhoods, such as
\emph{distance-$d$ neighborhoods}, are conceivable and are covered by
the theory in this section.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength graph, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

@@

When a vertex $i$ satisfies the generalized selection
condition~(\ref{eqn:general-selection}), no other valid $C$-point
assignments are able to affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then $i$
must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
$C$-point, regardless of other $C$-point selections and corresponding
weight updates made in the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes a
$C$-point, which is impossible until $i$ is assigned to the $C$-point
set.
\end{proof}

@@

Let $D_0$ be the set of all vertices satisfying
(\ref{eqn:general-selection}) following weight
initialization. Vertices in $D_0$ become $C$-points regardless of the
algorithm used to build coarse grids. In most independent set-based
algorithms developed so far, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

Lemma~\ref{5:lemma:D0_invariant} states that any coarse-grid selection
method using the general selection condition invariably selects $D_0$
vertices as $C$-points. This result is used to prove, given the
conditioned developed earlier in this section, the result of
coarse-grid selection is insensitive to the implementation decisions
made in specific algorithms.

\begin{thm}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{eqn:general-selection}), and weight
update heuristics as described above select identical coarse grids.
\end{thm}
\begin{proof}
  Let $c$ be the vertices in the extended selection neighborhood of
  $i$ (i.e., $ c \subset\mathcal{N}_i^{2s}$) satisfying the conditions
  for $D$ in some arbitrary iteration. Suppose
  assigning vertices $c_1 \subset c$ to the $C$-point set leads to
  $w_i > w_j$ for all $j \in \mathcal{N}_i^s$. Additionally, suppose
  assigning $c_2 \subset c$, $c_1 \ne c_2$, to the $C$-point set leads
  to $w_j > w_k$ for all $k \in \mathcal{N}_j^s$ for some $j \in
  \mathcal{N}_i^s$, $j \notin c$. In the first case, $i$ becomes a
  $C$-point before $j$. The opposite is true in the second case.

@@

Both cases are impossible, so the order of $C$-point selection within
the selection neighborhood of each vertex is invariant. Combined with
Lemma~\ref{5:lemma:D0_invariant} as the base case, induction proves
the invariance of coarse-grid selection for all algorithms using
identical selection and update conditions.
\end{proof}

@@

\section{Bucket-Sorted Independent Sets Selection}
\label{sec:bsis}
In this section, we introduce Bucket-Sorted Independent Sets
(BSIS). BSIS uses the same policies as the CLJP-c
algorithm~\cite{alber-cljpc}, so from the theory in
Section~\ref{sec:invariance}, BSIS and CLJP-c produce the same coarse
grids. BSIS is designed to be more efficient than CLJP-c.

\subsection{CLJP-c}
CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
$C$-points, where the colors are used as one component of the vertex
weights and the number of strong influences provides the other source
of contribution. As a result, the structure of the coarse grids
selected is improved over the CLJP algorithm~\cite{Cleary1998}. More
formally, the use of graph coloring provides the following important
result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
\label{cor:same-weight-indep-set}
Any set of vertices in the graph of $S$ sharing the same weight are an
independent set.
\end{cor}
\begin{cor}
\label{cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~(\ref{eqn:general-selection}). That is,
each vertex in that independent set has a uniquely maximal weight in
its neighborhood.
\end{cor}
Corollary~\ref{cor:same-weight-indep-set} states that independent sets
can be selected in the graph simply by selecting sets of vertices with
the same weight. Corollary~\ref{cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight in the
graph, updating weights, selecting the next set of vertices with
maximum weight, and so on. This is the approach taken by the BSIS
algorithm.

@@

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
%\textsc{Search-Graph}$(S,\, C,\, F)$%
$D \leftarrow \emptyset$
for $i \notin (C \cup F)$
  if $w_i > w_j\ \textrm{for all } j \in \mathcal{N}_i$
    $D \leftarrow D \cup \{i\}$
  end
end
return $D$
\end{lstlisting}
% \begin{algorithm}[h]
% \caption{Coarse-Grid Selection Graph Search}
% \label{5:alg:search}
% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
% \begin{algorithmic}[1]
% \STATE $D \leftarrow \emptyset$
% \FORALL{$i \notin (C \cup F)$}
% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
% \STATE $D \leftarrow D \cup \{i\}$
% \ENDIF
% \ENDFOR
% \RETURN $D$
% \end{algorithmic}
% \}
% \end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In Section~\ref{sec:bsis-alg},
we introduce the technique used in BSIS, which conducts the search
independent of the number of edges in the graph.

@@

Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
routines for search and weight update. Furthermore, rather than
applying the color information to augment vertex weights, BSIS uses
the colors in a bucket data structure. Once initialized, this data
structure selects independent sets, which satisfy the conditions in
(\ref{eq:D-selection}), in constant time.

@@

Timing data for the selection of all coarse grids in the hierarchy is
reported. This time includes the cost for all parts of the algorithms,
including the graph coloring phase. AMG solve phase information is not
reported since the algorithms produce identical coarse grids and
information on solve phase performance for AMG with CLJP-c is
documented in~\cite{alber-cljpc,alber-PCGS}.

The smallest problem is a $30 \times 30 \times 30$ grid. Subsequent
problems are grids of size $60^3$, $90^3$, and so on, up to
$210^3$. The largest problem is 343 times larger than the smallest
problem and contains more than nine million degrees of freedom.
